$p$-adic precision

### 2014-02-05 padicprec.tex
Split Section 3 in two Sections

\section{A new philosophy}
\label{sec:philosophy}

The aim of this section is to develop a new point of view on $p$-adic 
precision based on $p$-adic analysis. Actually, we shall generalize a 
bit the setting and work over a general ultrametric field (see definition
below).

@@

In this section, we recall usual definitions and properties related to
ultrametric analysis. We refer to Schneider \cite{Schneider} for a 
complete exposition.

@@

\subsection{The main Lemma}

@@

@@

Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic} have 
important applications to effective computations with $p$-adic numbers 
or power series, that we are going to discuss now. Let us recall first 
that a datum involving elements of a complete ultrametric field $K$ 
(\emph{e.g.} $p$-adic numbers or power series) cannot be stored entirely 
on a computed because they basically convey an infinite amount of 
informations. That is why, instead of this datum itself, we generally 
work with neighborhoods of it. In what follows, we shall consider 
neighborhoods of the form $x + H$ where $x$ is some element lying in a 
$K$-Banach $E$ and $H$ is a lattice in $E$. A very basic example is the 
approximate $p$-adic number $1 + O(p^N)$ which corresponds to the 
neighborhood of $1$ in $\Z_p$, which is the set of all $p$-adic numbers 
congruent to $1$ modulo $p^N$. We nevertheless emphasize that our 
definition is quite flexible and allows many neighborhoods that we do 
not consider in general. For instance, an ``admissible'' neighborhood 
around a monic polynomial $P_0 \in \Q_p[X]$ of degree $d$ may be defined 
by congruences of the form $P(a_i) \equiv P_0(a_i) \pmod {p^{n_i}}$ ($1 
\leq i \leq d$) for some fixed elements $a_i \in \Q_p$ and relative 
integers $n_i$.

Now, let us consider a function {\tt f} (in the sense of computer 
science) that takes as input an (approximate) element lying in an open 
subset $U$ of a $K$-Banach space $E$ and outputs another (approximate) 
element lying in an open subset $V$ of another $K$-Banach space $F$. Of 
course, this function can be modelised by a \emph{continuous} 
mathematical function $f : U \to V$. The relation is the following: when 
{\tt f} is called on the input $x + O(H)$, it outputs $x' + O(H')$ with 
$f(x+H) \subset x' + H'$. We say that {\tt f} does not generate 
artificial loss of precision if the above inclusion is an equality; it is 
often not the case as shown in \S \ref{ssec:stepbystep}.

@@

Many natural $p$-adic objects do not lie in vector spaces:
points in projective spaces or elliptic curves,
subspaces of a fixed vector space (which lie in Grassmannians),
classes of isomorphism of certain curves (which lie in various moduli spaces),
\emph{etc.}  In this section we extend the formalism 
developed in Section \ref{sec:philosophy} to a more general setting:
we consider the quite general case of 
differentiable manifolds locally modeled on ultrametric Banach spaces.
This covers all the 
aforementioned examples.

@@

For us, a QR factorization of a square matrix $M \in M_n(K)$ will be
a writing $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
orthogonal in the sense that $\trans Q \cdot Q$ is diagonal. As before, such 
a writing exists and is unique on a Zariski-open subset of $M_n(K)$. The
mapping $f : M \mapsto (Q,R)$ is then well defined on this subset. We 
would like to emphasize at this point that the orthogonality condition
defines a sub-manifold of $M_n(K)$ which is a apparently \emph{not} a
vector space (it is defined by equations of degree $2$). The codomain
of $f$ is then also a manifold; this example then fits to the setting
of \S \ref{sec:manifold} but not to those of \S \ref{sec:philosophy}. 
The generalization made in \S \ref{sec:manifold} is then needed.
Anyway, we can differentiate $f$ by following the method we used for LU 
factorization: differentiating the relation $M = QR$, we obtain:
\begin{equation}
\label{eq:diffQR}
\trans Q \cdot dM \cdot R^{-1} = \trans Q \cdot dQ + \Delta \cdot dR 
\cdot R^{-1}
\end{equation}
where $\Delta = \trans Q \cdot Q$ is a diagonal matrix by definition.
Moreover by differentiating $\trans Q \cdot Q = \Delta$, we find that
$\trans Q \cdot dQ$ can be written as the sum of an antisymmetric 
matrix and a diagonal one. Since moreove $dR \cdot R^{-1}$ is upper
triangular with all diagonal entries equal to $0$, we see that 
Eq.~\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
and $dM$.

>>>>

\section{The main Lemma}
\label{sec:mainlemma}

The theory of $p$-adic precision we want to develop is mainly based 
on a single lemma of $p$-adic analysis. The aim of this Section is
to state and prove this lemma. Consequences will be discussed after
in Section \ref{sec:tracking}.

From now, we generalize a bit the setting and work over a general 
ultrametric complete field (see definition below). This includes 
$p$-adic fields and Laurent series fields.

@@

We recall usual definitions and properties related to ultrametric 
analysis. We refer to Schneider \cite{Schneider} for a complete 
exposition.

@@

\subsection{Images of lattices under differentiable functions}

@@

\section{Tracking precision using differentials}
\label{sec:tracking}

Lemma \ref{lem:main} and Proposition \ref{prop:locanalytic} have 
important applications to effective computations with $p$-adic numbers 
or power series, that we discuss in this Section. Let us recall first 
that a datum involving elements of a complete ultrametric field $K$ 
(\emph{e.g.} $p$-adic numbers or power series) cannot be stored entirely 
on a computed because they basically convey an infinite amount of 
informations. That is why, instead of this datum itself, we generally 
work with neighborhoods of it.

In what follows, we shall consider neighborhoods of the form $x + H$ 
where $x$ is some element lying in a $K$-Banach $E$ and $H$ is a lattice 
in $E$. A very basic example is the approximate $p$-adic number $1 + 
O(p^N)$ which corresponds to the neighborhood of $1$ in $\Z_p$, which is 
the set of all $p$-adic numbers congruent to $1$ modulo $p^N$. We 
nevertheless emphasize that our definition is quite flexible and allows 
many neighborhoods that we do not consider in general. For instance, an 
``admissible'' neighborhood around a monic polynomial $P_0 \in \Q_p[X]$ 
of degree $d$ may be defined by congruences of the form $P(a_i) \equiv 
P_0(a_i) \pmod {p^{n_i}}$ ($1 \leq i \leq d$) for some fixed elements 
$a_i \in \Q_p$ and relative integers $n_i$.

@@

We consider a function {\tt f} (in the sense of computer 
science) that takes as input an (approximate) element lying in an open 
subset $U$ of a $K$-Banach space $E$ and outputs another (approximate) 
element lying in an open subset $V$ of another $K$-Banach space $F$. Of 
course, this function can be modelised by a \emph{continuous} 
mathematical function $f : U \to V$. The relation is the following: when 
{\tt f} is called on the input $x + O(H)$, it outputs $x' + O(H')$ with 
$f(x+H) \subset x' + H'$. We say that {\tt f} does not generate 
artificial loss of precision if the above inclusion is an equality; it is 
often not the case as shown in \S \ref{ssec:stepbystep}.

@@

Many natural $p$-adic objects do not lie in vector spaces:
points in projective spaces or elliptic curves,
subspaces of a fixed vector space (which lie in Grassmannians),
classes of isomorphism of certain curves (which lie in various moduli spaces),
\emph{etc.}  In this section we extend the formalism 
developed in Section \ref{sec:mainlemma} to a more general setting:
we consider the quite general case of 
differentiable manifolds locally modeled on ultrametric Banach spaces.
This covers all the 
aforementioned examples.

@@

For us, a QR factorization of a square matrix $M \in M_n(K)$ will be
a writing $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
orthogonal in the sense that $\trans Q \cdot Q$ is diagonal. As before, such 
a writing exists and is unique on a Zariski-open subset of $M_n(K)$. The
mapping $f : M \mapsto (Q,R)$ is then well defined on this subset. We 
would like to emphasize at this point that the orthogonality condition
defines a sub-manifold of $M_n(K)$ which is a apparently \emph{not} a
vector space (it is defined by equations of degree $2$). The codomain
of $f$ is then also a manifold; this example then fits to the setting
of \S \ref{sec:manifold} but not to those of \S \ref{sec:mainlemma}. 
The generalization made in \S \ref{sec:manifold} is then needed.
Anyway, we can differentiate $f$ by following the method we used for LU 
factorization: differentiating the relation $M = QR$, we obtain:
\begin{equation}
\label{eq:diffQR}
\trans Q \cdot dM \cdot R^{-1} = \trans Q \cdot dQ + \Delta \cdot dR 
\cdot R^{-1}
\end{equation}
where $\Delta = \trans Q \cdot Q$ is a diagonal matrix by definition.
Moreover by differentiating $\trans Q \cdot Q = \Delta$, we find that
$\trans Q \cdot dQ$ can be written as the sum of an antisymmetric 
matrix and a diagonal one. Since moreove $dR \cdot R^{-1}$ is upper
triangular with all diagonal entries equal to $0$, we see that 
Eq.~\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
and $dM$.
